---
title: "BN Structure Learning (Cross-Validated)"
author: "BPGbio Capstone"
date: "`r Sys.Date()`"
output: html_document
params:
  config: "config/L1b_clinical_curated.R"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE,
                      fig.width = 10, fig.height = 10)
knitr::opts_knit$set(root.dir = here::here("bn_cv"))
library(bnlearn)
library(dplyr)
library(ggplot2)
```

# 1. Load config and data

```{r load}
source(params$config)
cat(sprintf("Network: %s\n", NETWORK_ID))

bn_df <- readRDS(file.path("output", "models", sprintf("%s_data.rds", NETWORK_ID)))
cat(sprintf("Data: %d samples x %d nodes\n", nrow(bn_df), ncol(bn_df)))

set.seed(SEED)
```

# 2. Generate blacklist

No edges into root nodes + roots mutually independent.
CLG constraint: discrete nodes cannot have continuous parents.

```{r blacklist}
all_nodes <- names(bn_df)
non_roots <- setdiff(all_nodes, ROOTS)

# Rule 1: No edges INTO root nodes (from any node)
bl_into_roots <- expand.grid(from = all_nodes, to = ROOTS, stringsAsFactors = FALSE) %>%
  filter(from != to)

# Rule 2: Roots mutually independent (no edges between roots in either direction)
# Already covered by Rule 1 (roots are in the "to" set), but make explicit:
bl_between_roots <- expand.grid(from = ROOTS, to = ROOTS, stringsAsFactors = FALSE) %>%
  filter(from != to)

# Rule 3: Block edges between WBC and its deterministic components
# (WBC ≈ sum of the differential counts — same collinearity issue as the
# already-excluded NLR variable, but here we keep WBC as a node since it may
# carry non-redundant relationships to other markers, e.g. CRP/ESR).
# Filtered to intersect(all_nodes) since not every network config includes
# wbc or all 5 differential counts (e.g. L1b's curated panel omits basophils) —
# referencing a blacklist node that doesn't exist in bn_df throws an error in
# build.blacklist(), so this must stay defensive across all three configs.
wbc_components <- c("neutrophils", "lymphocytes", "monocytes", "eosinophils", "basophils")
wbc_cluster <- intersect(c("wbc", wbc_components), all_nodes)

if (length(wbc_cluster) >= 2) {
  bl_wbc <- expand.grid(from = wbc_cluster, to = wbc_cluster, stringsAsFactors = FALSE) %>%
    filter(from != to)
  cat(sprintf("WBC cluster present: %s (%d blacklist rows)\n",
              paste(wbc_cluster, collapse = ", "), nrow(bl_wbc)))
} else {
  bl_wbc <- data.frame(from = character(0), to = character(0))
  cat("WBC cluster: fewer than 2 relevant nodes present, no rule applied\n")
}

# Rule 4: Block direct parent-child edges in hierarchical modalities (e.g. freq's
# L1/L2/L3 cell-type gating tree). Only applies if the current config defines a
# PARENT_MAP (freq does; clinical/Olink/wb/pb don't) -- defensively checked with
# exists() so this safely no-ops for every other network, the same way the WBC
# rule's intersect() no-ops when wbc_cluster isn't relevant.
if (exists("PARENT_MAP")) {
  child_nodes <- intersect(names(PARENT_MAP), all_nodes)
  bl_hierarchy_rows <- lapply(child_nodes, function(child) {
    parent <- PARENT_MAP[[child]]
    if (parent %in% all_nodes) {
      data.frame(from = c(child, parent), to = c(parent, child), stringsAsFactors = FALSE)
    } else {
      NULL
    }
  })
  bl_hierarchy <- bind_rows(bl_hierarchy_rows)
  cat(sprintf("Hierarchy blacklist (PARENT_MAP): %d parent-child pairs, %d blacklist rows\n",
              nrow(bl_hierarchy) / 2, nrow(bl_hierarchy)))
} else {
  bl_hierarchy <- data.frame(from = character(0), to = character(0))
}

blacklist <- rbind(bl_into_roots, bl_between_roots, bl_wbc, bl_hierarchy) %>% distinct()
cat(sprintf("Blacklist: %d edges forbidden\n", nrow(blacklist)))
```

# 3. Score-based structure learning

## 3a. Hill-climbing

```{r hc}
set.seed(SEED)
bn_hc   <- hc(bn_df, score = "bic-cg", blacklist = blacklist, maxp = 10)
cat(sprintf("HC: %d edges, BIC-cg = %.2f\n", narcs(bn_hc), score(bn_hc, bn_df, type = "bic-cg")))
```

## 3b. Tabu search

```{r tabu}
set.seed(SEED)
bn_tabu <- tabu(bn_df, score = "bic-cg", blacklist = blacklist, maxp = 10)
cat(sprintf("Tabu: %d edges, BIC-cg = %.2f\n", narcs(bn_tabu), score(bn_tabu, bn_df, type = "bic-cg")))
```

## 3c. MMHC (hybrid)

```{r mmhc}
set.seed(SEED)
bn_mmhc <- mmhc(bn_df, blacklist = blacklist, maximize.args = list(maxp = 10))
cat(sprintf("MMHC: %d edges, BIC-cg = %.2f\n", narcs(bn_mmhc), score(bn_mmhc, bn_df, type = "bic-cg")))
```

# 4. Algorithm comparison

```{r algorithm-comparison}
# Edge overlap
arcs_hc   <- apply(arcs(bn_hc), 1, paste, collapse = "->")
arcs_tabu <- apply(arcs(bn_tabu), 1, paste, collapse = "->")
arcs_mmhc <- apply(arcs(bn_mmhc), 1, paste, collapse = "->")

all_arcs <- unique(c(arcs_hc, arcs_tabu, arcs_mmhc))
comparison <- data.frame(
  edge = all_arcs,
  hc   = all_arcs %in% arcs_hc,
  tabu = all_arcs %in% arcs_tabu,
  mmhc = all_arcs %in% arcs_mmhc,
  stringsAsFactors = FALSE
) %>%
  mutate(n_agree = hc + tabu + mmhc) %>%
  arrange(desc(n_agree), edge)

cat(sprintf("Total unique edges across algorithms: %d\n", nrow(comparison)))
cat(sprintf("Edges in all 3: %d\n", sum(comparison$n_agree == 3)))
cat(sprintf("Edges in 2/3: %d\n", sum(comparison$n_agree == 2)))
cat(sprintf("Edges in 1/3: %d\n", sum(comparison$n_agree == 1)))

knitr::kable(comparison, caption = "Edge agreement across algorithms")
```

# 5. Cross-validated structure learning

Replaces bootstrap resampling with repeated k-fold cross-validation, per sponsor
guidance. Bootstrap asked "is this edge stable under resampling?"; cross-validation
asks "does this structure generalize to held-out data?" — a direct, more
conventional overfitting check.

`bn.cv()` learns a fresh tabu structure on each training fold and scores it
against that fold's held-out data using conditional-Gaussian log-likelihood loss
(`"logl"`, bnlearn's log-likelihood loss — it uses each node's fitted CLG/discrete
distribution automatically based on the data's own node types, so no separate
CLG-specific loss label is needed here). `runs = 10` repeats the whole k-fold
procedure 10 times with different fold splits, for ~100 total model fits —
noticeably cheaper than the 1000-replicate bootstrap despite each individual
tabu search now covering more nodes.

**Note:** this chunk deliberately does NOT use `cache=TRUE` — a prior version of
this notebook cached the bootstrap chunk, which caused every network (regardless
of which config was passed in) to silently reuse the first cached result. Do not
reintroduce caching on this chunk without also keying it to `params$config`
(e.g. `cache.extra = tools::md5sum(params$config)`).

```{r cross-validation}
set.seed(SEED)
K_FOLDS <- 10
N_RUNS  <- 10

cv_result <- bn.cv(
  bn_df,
  bn = "tabu",
  algorithm.args = list(score = "bic-cg", blacklist = blacklist, maxp = 10),
  loss = "logl",
  method = "k-fold",
  k = K_FOLDS,
  runs = N_RUNS
)

# bn.cv() returns a list of N_RUNS "bn.kcv" objects when runs > 1 (a single
# "bn.kcv" object, itself a list of K_FOLDS folds, when runs == 1). Each fold
# element carries $fitted (a bn.fit learned on that fold's training data) and
# $loss (the held-out loss for that fold).
if (N_RUNS > 1) {
  fold_losses <- unlist(lapply(cv_result, function(run) sapply(run, function(fold) fold$loss)))
  cv_fits <- unlist(lapply(cv_result, function(run) lapply(run, function(fold) fold$fitted)),
                     recursive = FALSE)
} else {
  fold_losses <- sapply(cv_result, function(fold) fold$loss)
  cv_fits <- lapply(cv_result, function(fold) fold$fitted)
}

cat(sprintf("Cross-validation: %d folds x %d runs = %d total fits\n",
            K_FOLDS, N_RUNS, length(cv_fits)))
cat(sprintf("Held-out loss (logl): mean = %.3f, sd = %.3f\n",
            mean(fold_losses), sd(fold_losses)))

# Structure-only versions of each fold's fit, for custom.strength() below
cv_nets <- lapply(cv_fits, bn.net)
```

## 5a. Edge stability across CV folds

`custom.strength()` is the same underlying mechanism `boot.strength()` used —
generalized to accept any list of learned DAGs, not just bootstrap replicates.
Feeding it the CV-fold structures gives an edge-confidence table in the same
format as before, just derived from cross-validation instead of resampling.

```{r cv-strength}
cv_str <- custom.strength(cv_nets, nodes = names(bn_df))

cat("CV-derived strength summary:\n")
print(summary(cv_str$strength))
cat(sprintf("\nDirection probability summary:\n"))
print(summary(cv_str$direction))
```

## 5b. Averaged networks at multiple thresholds

```{r averaged-networks}
# Optimal threshold (bnlearn heuristic)
avg_opt <- averaged.network(cv_str)
thresh_opt <- attr(avg_opt, "threshold")
if (is.null(thresh_opt) || is.na(thresh_opt)) {
  cat("Optimal threshold heuristic returned NA/NULL — falling back to 0.5\n")
  thresh_opt <- 0.5
  avg_opt <- averaged.network(cv_str, threshold = thresh_opt)
}
cat(sprintf("Optimal threshold: %.4f\n", thresh_opt))
cat(sprintf("Averaged network (optimal, t=%.4f): %d edges\n", thresh_opt, narcs(avg_opt)))

# Conservative threshold — CV-appropriate empirical valley, not bootstrap's fixed 0.85
# (CV strengths cluster near 0/1 rather than spreading smoothly like bootstrap's,
# so a fixed 0.85 cutoff doesn't carry the same meaning across pipelines. Find the
# local minimum in the mid-range of the strength distribution instead, per-network.)
#
# Uses binned counts rather than density()/KDE: a smoothed density estimate gets
# pulled toward whichever boundary is closest to the near-1.0 cluster (confirmed
# on L1a, where KDE reported 0.95 — right at the search boundary — while the
# real minimum-count bin was 0.80-0.85). Binning can't overshoot like that.
strengths_mid <- cv_str$strength[cv_str$strength > 0.3 & cv_str$strength < 0.95]
if (length(strengths_mid) >= 5) {
  bin_width <- 0.05
  breaks <- seq(0.3, 0.95, by = bin_width)
  h <- hist(strengths_mid, breaks = breaks, plot = FALSE)
  min_bin <- which.min(h$counts)
  thresh_conservative <- h$mids[min_bin]
  cat(sprintf("Bin counts (0.3-0.95, width %.2f): %s\n",
              bin_width, paste(h$counts, collapse = ", ")))
} else {
  cat("Too few mid-range edge strengths to estimate a valley — falling back to 0.85\n")
  thresh_conservative <- 0.85
}
cat(sprintf("CV-appropriate conservative threshold (empirical valley): %.3f\n", thresh_conservative))

avg_conservative <- averaged.network(cv_str, threshold = thresh_conservative)
cat(sprintf("Averaged network (conservative, t=%.4f): %d edges\n", thresh_conservative, narcs(avg_conservative)))

# Liberal threshold (0.50)
avg_50 <- averaged.network(cv_str, threshold = 0.50)
cat(sprintf("Averaged network (t=0.50): %d edges\n", narcs(avg_50)))
```

# 6. Edge strength table

```{r edge-strength}
# Filter to edges with strength > 0.1 for readability
sig_edges <- cv_str %>%
  filter(strength > 0.1) %>%
  arrange(desc(strength))

knitr::kable(sig_edges, digits = 3,
             caption = "Cross-validated edge strengths (strength > 0.1)")
```

# 7. DAG visualizations

```{r plot-dag-tabu, fig.cap="Tabu search DAG"}
if (requireNamespace("Rgraphviz", quietly = TRUE)) {
  graphviz.plot(bn_tabu, main = sprintf("%s — Tabu", NETWORK_ID),
                shape = "ellipse",
                highlight = list(nodes = ROOTS, fill = "lightblue"))
} else {
  # Fallback: igraph/ggraph
  library(igraph)
  library(ggraph)
  g <- igraph::graph_from_data_frame(arcs(bn_tabu))
  ggraph(g, layout = "fr") +
    geom_edge_link(arrow = arrow(length = unit(3, "mm")), end_cap = circle(3, "mm")) +
    geom_node_point(size = 5, aes(color = name %in% ROOTS)) +
    geom_node_text(aes(label = name), repel = TRUE, size = 3) +
    scale_color_manual(values = c("FALSE" = "grey70", "TRUE" = "lightblue"),
                       guide = "none") +
    labs(title = sprintf("%s — Tabu", NETWORK_ID)) +
    theme_void()
}
```

```{r plot-dag-avg, fig.cap="Averaged network (optimal threshold)"}
if (requireNamespace("Rgraphviz", quietly = TRUE)) {
  graphviz.plot(avg_opt, main = sprintf("%s — CV Averaged (t=%.3f)", NETWORK_ID, thresh_opt),
                shape = "ellipse",
                highlight = list(nodes = ROOTS, fill = "lightblue"))
} else {
  library(igraph)
  library(ggraph)
  g <- igraph::graph_from_data_frame(arcs(avg_opt))
  ggraph(g, layout = "fr") +
    geom_edge_link(arrow = arrow(length = unit(3, "mm")), end_cap = circle(3, "mm")) +
    geom_node_point(size = 5, aes(color = name %in% ROOTS)) +
    geom_node_text(aes(label = name), repel = TRUE, size = 3) +
    scale_color_manual(values = c("FALSE" = "grey70", "TRUE" = "lightblue"),
                       guide = "none") +
    labs(title = sprintf("%s — CV Averaged (t=%.3f)", NETWORK_ID, thresh_opt)) +
    theme_void()
}
```

# 8. Fit parameters and save

```{r fit-and-save}
# Fit parameters on the tabu DAG (learned on the FULL data, not a CV fold)
fit_tabu <- bn.fit(bn_tabu, bn_df)

# Save everything
models <- list(
  bn_hc            = bn_hc,
  bn_tabu          = bn_tabu,
  bn_mmhc          = bn_mmhc,
  fit_tabu         = fit_tabu,
  cv_str           = cv_str,
  cv_losses        = fold_losses,
  avg_opt          = avg_opt,
  avg_conservative = avg_conservative,
  avg_50           = avg_50,
  blacklist        = blacklist,
  config           = list(
    NETWORK_ID       = NETWORK_ID,
    SEED             = SEED,
    ROOTS            = ROOTS,
    NODE_MAP         = NODE_MAP,
    LOG_TRANSFORM    = LOG_TRANSFORM,
    REFERENCE_LEVELS = REFERENCE_LEVELS,
    K_FOLDS          = K_FOLDS,
    N_RUNS           = N_RUNS
  )
)

out_path <- file.path("output", "models", sprintf("%s_models.rds", NETWORK_ID))
saveRDS(models, out_path)
cat(sprintf("Saved all models: %s\n", out_path))
```

# 9. Save DAG figures

```{r save-figures}
fig_dir <- file.path("output", "figures")

# Edge strength histogram
p <- ggplot(cv_str, aes(x = strength)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = c(0.5, 0.85, thresh_opt),
             linetype = c("dashed", "dashed", "solid"),
             color = c("orange", "red", "darkgreen")) +
  annotate("text", x = 0.5, y = Inf, label = "0.50", vjust = 2, color = "orange") +
  annotate("text", x = 0.85, y = Inf, label = "0.85", vjust = 2, color = "red") +
  annotate("text", x = thresh_opt, y = Inf, label = sprintf("opt=%.2f", thresh_opt),
           vjust = 3.5, color = "darkgreen") +
  labs(title = sprintf("%s — CV Edge Strength Distribution", NETWORK_ID),
       x = "Edge Strength", y = "Count") +
  theme_minimal()

ggsave(file.path(fig_dir, sprintf("%s_edge_strength_hist.png", NETWORK_ID)),
       p, width = 8, height = 5, dpi = 150)
cat("Saved edge strength histogram\n")

# Held-out loss distribution — new, CV-specific diagnostic with no bootstrap
# equivalent. Shows whether generalization performance is stable across folds
# or highly variable (a high-variance loss suggests the model is sensitive to
# exactly which subjects end up in the training set — a red flag given n~90).
p_loss <- ggplot(data.frame(loss = fold_losses), aes(x = loss)) +
  geom_histogram(bins = 20, fill = "darkorange", color = "white") +
  labs(title = sprintf("%s — CV Held-Out Loss Distribution (logl)", NETWORK_ID),
       x = "Held-out loss (lower = better fit)", y = "Count") +
  theme_minimal()

ggsave(file.path(fig_dir, sprintf("%s_cv_loss_hist.png", NETWORK_ID)),
       p_loss, width = 8, height = 5, dpi = 150)
cat("Saved CV loss histogram\n")
```
