---
title: "BN Structure Learning — All Variables (Cross-Validated)"
author: "BPGbio Capstone"
date: "`r Sys.Date()`"
output: html_document
params:
  config: "config/L_all.R"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE)
knitr::opts_knit$set(root.dir = "/Users/shaveen/Desktop/bn_cv")
library(bnlearn)
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
```

# 1. Load config and prepared data

Unlike every other network's structure-learning file, this one does NOT
re-derive `bn_df` from `CONTINUOUS_MAP` (L_all's config doesn't define one --
node selection is programmatic, built by `01_data_prep_all_cv.Rmd`). Instead
it loads the already-prepared `bn_complete` that file exported. **Run
`01_data_prep_all_cv.Rmd` first** -- this file will fail to find its input
otherwise.

```{r load-config}
source(params$config)
if (!exists("MAXP")) {
  MAXP <- 10L
  cat("Config did not define MAXP -- falling back to default of 10.\n")
}
cat(sprintf("Network: %s (seed=%d, maxp=%d)\n", NETWORK_ID, SEED, MAXP))

data_path <- file.path("output", "models", sprintf("%s_data.rds", NETWORK_ID))
if (!file.exists(data_path)) {
  stop(sprintf("Prepared data not found at %s -- run 01_data_prep_all_cv.Rmd first.", data_path))
}
bn_df <- readRDS(data_path)
cat(sprintf("Loaded prepared data: %d samples x %d nodes\n", nrow(bn_df), ncol(bn_df)))

mapping_path <- file.path("output", sprintf("%s_node_mapping.csv", NETWORK_ID))
if (!file.exists(mapping_path)) {
  stop(sprintf("Node mapping not found at %s -- run 01_data_prep_all_cv.Rmd first.", mapping_path))
}
node_mapping <- read_csv(mapping_path, show_col_types = FALSE)
```

# 2. Generate blacklist

No edges into root nodes + roots mutually independent. CLG constraint:
discrete nodes cannot have continuous parents.

```{r blacklist}
all_nodes <- names(bn_df)
non_roots <- setdiff(all_nodes, ROOTS)

# Rule 1: No edges INTO root nodes (from any node)
bl_into_roots <- expand.grid(from = all_nodes, to = ROOTS, stringsAsFactors = FALSE) %>%
  filter(from != to)

# Rule 2: Roots mutually independent (no edges between roots in either direction)
bl_between_roots <- expand.grid(from = ROOTS, to = ROOTS, stringsAsFactors = FALSE) %>%
  filter(from != to)

# Rule 3: Block edges between WBC and its deterministic components.
# IMPORTANT: this network's aliases are generated programmatically by
# 01_data_prep_all_cv.Rmd (e.g. clinical.bc.wbc -> "bc_wbc"), NOT the short
# hand-picked names ("wbc", "neutrophils", ...) used in L1a/L1b's manually-
# written configs. Matching on hardcoded alias strings here would silently
# find zero nodes and no-op -- the redundancy would still be present in the
# data, just undetected. Instead, look up the cluster by its actual CSV
# column identity via node_mapping.csv, which is correct regardless of
# which alias each column ended up with.
wbc_csv_targets <- c(
  "clinical.bc.wbc",
  "clinical.bc.neutrophil_count",
  "clinical.bc.lymphocyte_count",
  "clinical.bc.monocyte_count",
  "clinical.bc.eosinophil_count",
  "clinical.bc.basophil_count"
)
wbc_cluster <- node_mapping$alias[node_mapping$csv_column %in% wbc_csv_targets]
wbc_cluster <- intersect(wbc_cluster, all_nodes)

if (length(wbc_cluster) >= 2) {
  bl_wbc <- expand.grid(from = wbc_cluster, to = wbc_cluster, stringsAsFactors = FALSE) %>%
    filter(from != to)
  cat(sprintf("WBC cluster present: %s (%d blacklist rows)\n",
              paste(wbc_cluster, collapse = ", "), nrow(bl_wbc)))
} else {
  bl_wbc <- data.frame(from = character(0), to = character(0))
  cat("WBC cluster: fewer than 2 relevant nodes present, no rule applied\n")
}

# Rule 4: Block direct parent-child edges in hierarchical modalities (freq's
# L1/L2/L3 cell-type gating tree). Freq's programmatic aliases already match
# PARENT_MAP's key format directly (both use lowercase l1_/l2_/l3_ prefixes),
# so this can use PARENT_MAP as-is, unlike the WBC rule above.
if (exists("PARENT_MAP")) {
  child_nodes <- intersect(names(PARENT_MAP), all_nodes)
  bl_hierarchy_rows <- lapply(child_nodes, function(child) {
    parent <- PARENT_MAP[[child]]
    if (parent %in% all_nodes) {
      data.frame(from = c(child, parent), to = c(parent, child), stringsAsFactors = FALSE)
    } else {
      NULL
    }
  })
  bl_hierarchy <- bind_rows(bl_hierarchy_rows)
  cat(sprintf("Hierarchy blacklist (PARENT_MAP): %d parent-child pairs, %d blacklist rows\n",
              nrow(bl_hierarchy) / 2, nrow(bl_hierarchy)))
} else {
  bl_hierarchy <- data.frame(from = character(0), to = character(0))
}

blacklist <- rbind(bl_into_roots, bl_between_roots, bl_wbc, bl_hierarchy) %>% distinct()
cat(sprintf("Blacklist: %d edges forbidden\n", nrow(blacklist)))
```

# 3. Score-based structure learning

## 3a. Hill-climbing

```{r hc}
set.seed(SEED)
bn_hc <- hc(bn_df, score = "bic-cg", blacklist = blacklist, maxp = MAXP)
cat(sprintf("HC: %d edges, BIC-cg = %.2f\n", narcs(bn_hc), score(bn_hc, bn_df, type = "bic-cg")))
```

## 3b. Tabu search

```{r tabu}
set.seed(SEED)
bn_tabu <- tabu(bn_df, score = "bic-cg", blacklist = blacklist, maxp = MAXP)
cat(sprintf("Tabu: %d edges, BIC-cg = %.2f\n", narcs(bn_tabu), score(bn_tabu, bn_df, type = "bic-cg")))
```

## 3c. MMHC (hybrid)

```{r mmhc}
set.seed(SEED)
bn_mmhc <- mmhc(bn_df, blacklist = blacklist, maximize.args = list(maxp = MAXP))
cat(sprintf("MMHC: %d edges, BIC-cg = %.2f\n", narcs(bn_mmhc), score(bn_mmhc, bn_df, type = "bic-cg")))
```

# 4. Algorithm comparison

```{r algo-comparison}
arcs_hc <- apply(arcs(bn_hc), 1, paste, collapse = "->")
arcs_tabu <- apply(arcs(bn_tabu), 1, paste, collapse = "->")
arcs_mmhc <- apply(arcs(bn_mmhc), 1, paste, collapse = "->")

cat(sprintf("HC-Tabu overlap: %d / %d\n", length(intersect(arcs_hc, arcs_tabu)), length(union(arcs_hc, arcs_tabu))))
cat(sprintf("Tabu-MMHC overlap: %d / %d\n", length(intersect(arcs_tabu, arcs_mmhc)), length(union(arcs_tabu, arcs_mmhc))))
cat(sprintf("All three agree on: %d edges\n", length(Reduce(intersect, list(arcs_hc, arcs_tabu, arcs_mmhc)))))
```

# 5. Cross-validation

`bn.cv()` learns a fresh tabu structure on each training fold and scores it
against the held-out fold.

```{r cross-validation}
K_FOLDS <- 10
N_RUNS  <- 10

cv_result <- bn.cv(
  bn_df,
  bn = "tabu",
  algorithm.args = list(score = "bic-cg", blacklist = blacklist, maxp = MAXP),
  loss = "logl",
  method = "k-fold",
  k = K_FOLDS,
  runs = N_RUNS
)

fold_losses <- unlist(lapply(cv_result, function(run) sapply(run, function(fold) fold$loss)))
cat(sprintf("Held-out loss (logl): mean = %.3f, sd = %.3f\n", mean(fold_losses), sd(fold_losses)))

cv_fits <- unlist(lapply(cv_result, function(run) lapply(run, function(fold) fold$fitted)), recursive = FALSE)
cv_nets <- lapply(cv_fits, bn.net)
```

## 5a. Edge stability across CV folds

```{r cv-strength}
cv_str <- custom.strength(cv_nets, nodes = names(bn_df))
```

## 5b. Averaged networks at multiple thresholds

```{r averaged-networks}
# Optimal threshold (bnlearn heuristic)
avg_opt <- averaged.network(cv_str)
thresh_opt <- attr(avg_opt, "threshold")
if (is.null(thresh_opt) || is.na(thresh_opt)) {
  cat("Optimal threshold heuristic returned NA/NULL -- falling back to 0.5\n")
  thresh_opt <- 0.5
  avg_opt <- averaged.network(cv_str, threshold = thresh_opt)
}
cat(sprintf("Optimal threshold: %.4f\n", thresh_opt))
cat(sprintf("Averaged network (optimal, t=%.4f): %d edges\n", thresh_opt, narcs(avg_opt)))

# Conservative threshold -- CV-appropriate empirical valley, not bootstrap's fixed 0.85
strengths_mid <- cv_str$strength[cv_str$strength > 0.3 & cv_str$strength < 0.95]
if (length(strengths_mid) >= 5) {
  bin_width <- 0.05
  breaks <- seq(0.3, 0.95, by = bin_width)
  h <- hist(strengths_mid, breaks = breaks, plot = FALSE)
  min_bin <- which.min(h$counts)
  thresh_conservative <- h$mids[min_bin]
  cat(sprintf("Bin counts (0.3-0.95, width %.2f): %s\n",
              bin_width, paste(h$counts, collapse = ", ")))
} else {
  cat("Too few mid-range edge strengths to estimate a valley -- falling back to 0.85\n")
  thresh_conservative <- 0.85
}
cat(sprintf("CV-appropriate conservative threshold (empirical valley): %.3f\n", thresh_conservative))

avg_conservative <- averaged.network(cv_str, threshold = thresh_conservative)
cat(sprintf("Averaged network (conservative, t=%.4f): %d edges\n", thresh_conservative, narcs(avg_conservative)))

# Liberal threshold (0.50)
avg_50 <- averaged.network(cv_str, threshold = 0.50)
cat(sprintf("Averaged network (t=0.50): %d edges\n", narcs(avg_50)))
```

# 6. Edge strength table

```{r edge-table}
# Filter to edges with strength > 0.1 for readability
sig_edges <- cv_str %>% filter(strength > 0.1) %>% arrange(desc(strength))
knitr::kable(head(sig_edges, 50), digits = 3)

# Full export -- every directed pair, unfiltered (see README for column
# meanings: strength/direction match the histogram/table above,
# in_final_dag marks membership in avg_conservative specifically)
edges_export <- cv_str %>%
  mutate(in_final_dag = mapply(function(f, t) {
    arcs_conservative <- arcs(avg_conservative)
    any(arcs_conservative[, "from"] == f & arcs_conservative[, "to"] == t)
  }, from, to))
write_csv(edges_export, file.path("output", sprintf("%s_literature_edges_cv.csv", NETWORK_ID)))
cat(sprintf("Saved edge export: output/%s_literature_edges_cv.csv (%d rows)\n",
            NETWORK_ID, nrow(edges_export)))
```

# 7. Diagnostic plots

```{r edge-strength-histogram}
fig_dir <- "output/figures"
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)

p_strength <- ggplot(cv_str, aes(x = strength)) +
  geom_histogram(bins = 30, fill = "steelblue") +
  geom_vline(xintercept = 0.5, color = "darkgreen") +
  geom_vline(xintercept = 0.85, color = "red", linetype = "dashed") +
  annotate("text", x = 0.5, y = Inf, label = sprintf("0.50\nopt=%.2f", thresh_opt),
           vjust = 1.5, color = "darkgreen", size = 3) +
  annotate("text", x = 0.85, y = Inf, label = "0.85", vjust = 1.5, color = "red", size = 3) +
  labs(title = sprintf("%s — CV Edge Strength Distribution", NETWORK_ID),
       x = "Edge Strength", y = "Count") +
  theme_minimal()
ggsave(file.path(fig_dir, sprintf("%s_edge_strength_hist_cv.png", NETWORK_ID)),
       p_strength, width = 8, height = 5, dpi = 150)

# Held-out loss distribution
loss_df <- data.frame(loss = fold_losses)
p_loss <- ggplot(loss_df, aes(x = loss)) +
  geom_histogram(bins = 30, fill = "orange") +
  labs(title = sprintf("%s — CV Held-Out Loss Distribution (logl)", NETWORK_ID),
       x = "Held-out loss (lower = better fit)", y = "Count") +
  theme_minimal()
ggsave(file.path(fig_dir, sprintf("%s_loss_hist_cv.png", NETWORK_ID)),
       p_loss, width = 8, height = 5, dpi = 150)

cat("Saved diagnostic plots to output/figures/\n")
```

```{r dag-visualizations}
net_dir <- "Networks"
if (!dir.exists(net_dir)) dir.create(net_dir, recursive = TRUE)

png(file.path(net_dir, sprintf("%s_tabu_cv.png", NETWORK_ID)), width = 1200, height = 1200)
plot(bn_tabu, main = sprintf("%s — Tabu", NETWORK_ID))
dev.off()

png(file.path(net_dir, sprintf("%s_avg_cv.png", NETWORK_ID)), width = 1200, height = 1200)
plot(avg_opt, main = sprintf("%s — CV Averaged (t=%.3f)", NETWORK_ID, thresh_opt))
dev.off()

cat("Saved DAG visualizations to Networks/\n")
```

# 8. Fit parameters and save

```{r fit-and-save}
# Fit parameters on the tabu DAG (learned on the FULL data, not a CV fold)
fit_tabu <- bn.fit(bn_tabu, bn_df)

models <- list(
  bn_hc            = bn_hc,
  bn_tabu          = bn_tabu,
  bn_mmhc          = bn_mmhc,
  fit_tabu         = fit_tabu,
  cv_str           = cv_str,
  cv_losses        = fold_losses,
  avg_opt          = avg_opt,
  avg_conservative = avg_conservative,
  avg_50           = avg_50,
  blacklist        = blacklist,
  maxp_used        = MAXP,
  config           = list(
    network_id = NETWORK_ID,
    seed       = SEED
  )
)

out_path <- file.path("output", "models", sprintf("%s_models.rds", NETWORK_ID))
saveRDS(models, out_path)
cat(sprintf("Saved: %s\n", out_path))
```
