---
title: "BN Structure Learning"
author: "BPGbio Capstone"
date: "`r Sys.Date()`"
output: html_document
params:
  config: "config/L1b_clinical_curated.R"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE,
                      fig.width = 10, fig.height = 10)
knitr::opts_knit$set(root.dir = here::here("bn_learning"))
library(bnlearn)
library(dplyr)
library(ggplot2)
```

# 1. Load config and data

```{r load}
source(params$config)
cat(sprintf("Network: %s\n", NETWORK_ID))

bn_df <- readRDS(file.path("output", "models", sprintf("%s_data.rds", NETWORK_ID)))

# bnlearn's C backend does NOT support tibbles. A tibble keeps its tbl_df class
# through subsetting, which makes bn.fit misread factor columns and throw
# "REAL() can only be applied to a 'numeric', not a 'integer'" on conditional-
# Gaussian nodes. Coerce to a plain data.frame before any bnlearn call.
bn_df <- as.data.frame(bn_df)

# Belt-and-suspenders: ensure continuous columns are double, not integer.
for (col in setdiff(names(bn_df), ROOTS)) {
  bn_df[[col]] <- as.numeric(bn_df[[col]])
}

cat(sprintf("Data: %d samples x %d nodes\n", nrow(bn_df), ncol(bn_df)))

set.seed(SEED)
MAXP <- if (exists("MAXP", inherits = FALSE)) MAXP else 10L
cat(sprintf("maxp: %d\n", MAXP))
```

# 2. Generate blacklist

No edges into root nodes + roots mutually independent.
CLG constraint: discrete nodes cannot have continuous parents.

```{r blacklist}
all_nodes <- names(bn_df)
non_roots <- setdiff(all_nodes, ROOTS)

# Rule 1: No edges INTO root nodes (from any node)
bl_into_roots <- expand.grid(from = all_nodes, to = ROOTS, stringsAsFactors = FALSE) %>%
  filter(from != to)

# Rule 2: Roots mutually independent (no edges between roots in either direction)
# Already covered by Rule 1 (roots are in the "to" set), but make explicit:
bl_between_roots <- expand.grid(from = ROOTS, to = ROOTS, stringsAsFactors = FALSE) %>%
  filter(from != to)

# Rule 3: Block edges between WBC and its deterministic components
# (WBC ≈ sum of the differential counts — same collinearity issue as the
# already-excluded NLR variable, but here we keep WBC as a node since it may
# carry non-redundant relationships to other markers, e.g. CRP/ESR).
# Filtered to intersect(all_nodes) since not every network config includes
# wbc or all 5 differential counts (e.g. L1b's curated panel omits basophils) —
# referencing a blacklist node that doesn't exist in bn_df throws an error in
# build.blacklist(), so this must stay defensive across all three configs.
wbc_components <- c("neutrophils", "lymphocytes", "monocytes", "eosinophils", "basophils")
wbc_cluster <- intersect(c("wbc", wbc_components), all_nodes)

if (length(wbc_cluster) >= 2) {
  bl_wbc <- expand.grid(from = wbc_cluster, to = wbc_cluster, stringsAsFactors = FALSE) %>%
    filter(from != to)
  cat(sprintf("WBC cluster present: %s (%d blacklist rows)\n",
              paste(wbc_cluster, collapse = ", "), nrow(bl_wbc)))
} else {
  bl_wbc <- data.frame(from = character(0), to = character(0))
  cat("WBC cluster: fewer than 2 relevant nodes present, no rule applied\n")
}

# Rule 4: Block direct parent-child edges in hierarchical modalities (e.g. freq's
# L1/L2/L3 cell-type gating tree). Only applies if the current config defines a
# PARENT_MAP (freq does; clinical/Olink/wb/pb don't) -- defensively checked with
# exists() so this safely no-ops for every other network, the same way the WBC
# rule's intersect() no-ops when wbc_cluster isn't relevant.
if (exists("PARENT_MAP")) {
  child_nodes <- intersect(names(PARENT_MAP), all_nodes)
  bl_hierarchy_rows <- lapply(child_nodes, function(child) {
    parent <- PARENT_MAP[[child]]
    if (parent %in% all_nodes) {
      data.frame(from = c(child, parent), to = c(parent, child), stringsAsFactors = FALSE)
    } else {
      NULL
    }
  })
  bl_hierarchy <- bind_rows(bl_hierarchy_rows)
  cat(sprintf("Hierarchy blacklist (PARENT_MAP): %d parent-child pairs, %d blacklist rows\n",
              nrow(bl_hierarchy) / 2, nrow(bl_hierarchy)))
} else {
  bl_hierarchy <- data.frame(from = character(0), to = character(0))
}

blacklist <- rbind(bl_into_roots, bl_between_roots, bl_wbc, bl_hierarchy) %>% distinct()
cat(sprintf("Blacklist: %d edges forbidden\n", nrow(blacklist)))
```

# 3. Score-based structure learning

## 3a. Hill-climbing

```{r hc}
set.seed(SEED)
bn_hc <- hc(bn_df, score = "bic-cg", blacklist = blacklist, maxp = MAXP)
cat(sprintf("HC: %d edges, BIC-cg = %.2f\n", narcs(bn_hc), score(bn_hc, bn_df, type = "bic-cg")))
```

## 3b. Tabu search

```{r tabu}
set.seed(SEED)
bn_tabu <- tabu(bn_df, score = "bic-cg", blacklist = blacklist, maxp = MAXP)
cat(sprintf("Tabu: %d edges, BIC-cg = %.2f\n", narcs(bn_tabu), score(bn_tabu, bn_df, type = "bic-cg")))
```

## 3c. MMHC (hybrid)

```{r mmhc}
set.seed(SEED)
bn_mmhc <- mmhc(bn_df, blacklist = blacklist, maximize.args = list(maxp = MAXP))
cat(sprintf("MMHC: %d edges, BIC-cg = %.2f\n", narcs(bn_mmhc), score(bn_mmhc, bn_df, type = "bic-cg")))
```

# 4. Algorithm comparison

```{r algorithm-comparison}
# Edge overlap
arcs_hc   <- apply(arcs(bn_hc), 1, paste, collapse = "->")
arcs_tabu <- apply(arcs(bn_tabu), 1, paste, collapse = "->")
arcs_mmhc <- apply(arcs(bn_mmhc), 1, paste, collapse = "->")

all_arcs <- unique(c(arcs_hc, arcs_tabu, arcs_mmhc))
comparison <- data.frame(
  edge = all_arcs,
  hc   = all_arcs %in% arcs_hc,
  tabu = all_arcs %in% arcs_tabu,
  mmhc = all_arcs %in% arcs_mmhc,
  stringsAsFactors = FALSE
) %>%
  mutate(n_agree = hc + tabu + mmhc) %>%
  arrange(desc(n_agree), edge)

cat(sprintf("Total unique edges across algorithms: %d\n", nrow(comparison)))
cat(sprintf("Edges in all 3: %d\n", sum(comparison$n_agree == 3)))
cat(sprintf("Edges in 2/3: %d\n", sum(comparison$n_agree == 2)))
cat(sprintf("Edges in 1/3: %d\n", sum(comparison$n_agree == 1)))

knitr::kable(comparison, caption = "Edge agreement across algorithms")
```

# 5. Bootstrap model averaging

```{r bootstrap}
# NOTE: do NOT add cache=TRUE here. knitr's cache key is based on chunk code,
# not params$config, so cached results from one network silently reuse for all
# subsequent networks. If caching is ever needed, key it to the config file:
# cache.extra = tools::md5sum(params$config)
set.seed(SEED)
boot_str <- boot.strength(
  bn_df,
  R = 1000,
  algorithm = "tabu",
  algorithm.args = list(score = "bic-cg", blacklist = blacklist, maxp = MAXP)
)

cat("Bootstrap strength summary:\n")
print(summary(boot_str$strength))
cat(sprintf("\nDirection probability summary:\n"))
print(summary(boot_str$direction))
```

## 5a. Averaged networks at multiple thresholds

```{r averaged-networks}
# Optimal threshold (bnlearn heuristic)
avg_opt <- averaged.network(boot_str)
thresh_opt <- attr(avg_opt, "threshold")
if (is.null(thresh_opt) || is.na(thresh_opt)) {
  cat("Optimal threshold heuristic returned NA/NULL — falling back to 0.5\n")
  thresh_opt <- 0.5
  avg_opt <- averaged.network(boot_str, threshold = thresh_opt)
}
cat(sprintf("Optimal threshold: %.4f\n", thresh_opt))
cat(sprintf("Averaged network (optimal, t=%.4f): %d edges\n", thresh_opt, narcs(avg_opt)))

# Conservative threshold — empirical valley in mid-range strengths (0.3-0.95).
# Uses binned histogram rather than density()/KDE: a smoothed density estimate
# can be pulled toward whichever boundary is closest to the near-1.0 cluster
# (confirmed on L1a, where KDE reported 0.95 — right at the search boundary —
# while the real minimum-count bin was 0.80-0.85). Binning can't overshoot.
# Fallback to 0.85 when too few mid-range edges exist to estimate a valley.
strengths_mid <- boot_str$strength[boot_str$strength > 0.3 & boot_str$strength < 0.95]
if (length(strengths_mid) >= 5) {
  bin_width <- 0.05
  breaks <- seq(0.3, 0.95, by = bin_width)
  h <- hist(strengths_mid, breaks = breaks, plot = FALSE)
  min_bin <- which.min(h$counts)
  thresh_conservative <- h$mids[min_bin]
  cat(sprintf("Bin counts (0.3-0.95, width %.2f): %s\n",
              bin_width, paste(h$counts, collapse = ", ")))
} else {
  cat("Too few mid-range edge strengths to estimate a valley — falling back to 0.85\n")
  thresh_conservative <- 0.85
}
cat(sprintf("Conservative threshold (empirical valley): %.3f\n", thresh_conservative))

avg_conservative <- averaged.network(boot_str, threshold = thresh_conservative)
cat(sprintf("Averaged network (conservative, t=%.4f): %d edges\n", thresh_conservative, narcs(avg_conservative)))

# Liberal threshold (0.50)
avg_50 <- averaged.network(boot_str, threshold = 0.50)
cat(sprintf("Averaged network (t=0.50): %d edges\n", narcs(avg_50)))
```

# 6. Edge strength table

```{r edge-strength}
# Filter to edges with strength > 0.1 for readability
sig_edges <- boot_str %>%
  filter(strength > 0.1) %>%
  arrange(desc(strength))

knitr::kable(sig_edges, digits = 3,
             caption = "Bootstrap edge strengths (strength > 0.1)")
```

# 7. DAG visualizations

```{r plot-dag-tabu, fig.cap="Tabu search DAG"}
if (requireNamespace("Rgraphviz", quietly = TRUE)) {
  graphviz.plot(bn_tabu, main = sprintf("%s — Tabu", NETWORK_ID),
                shape = "ellipse",
                highlight = list(nodes = ROOTS, fill = "lightblue"))
} else {
  # Fallback: igraph/ggraph
  library(igraph)
  library(ggraph)
  g <- igraph::graph_from_data_frame(arcs(bn_tabu))
  ggraph(g, layout = "fr") +
    geom_edge_link(arrow = arrow(length = unit(3, "mm")), end_cap = circle(3, "mm")) +
    geom_node_point(size = 5, aes(color = name %in% ROOTS)) +
    geom_node_text(aes(label = name), repel = TRUE, size = 3) +
    scale_color_manual(values = c("FALSE" = "grey70", "TRUE" = "lightblue"),
                       guide = "none") +
    labs(title = sprintf("%s — Tabu", NETWORK_ID)) +
    theme_void()
}
```

```{r plot-dag-avg, fig.cap="Averaged network (optimal threshold)"}
if (requireNamespace("Rgraphviz", quietly = TRUE)) {
  graphviz.plot(avg_opt, main = sprintf("%s — Averaged (t=%.3f)", NETWORK_ID, as.numeric(thresh_opt)),
                shape = "ellipse",
                highlight = list(nodes = ROOTS, fill = "lightblue"))
} else {
  library(igraph)
  library(ggraph)
  g <- igraph::graph_from_data_frame(arcs(avg_opt))
  ggraph(g, layout = "fr") +
    geom_edge_link(arrow = arrow(length = unit(3, "mm")), end_cap = circle(3, "mm")) +
    geom_node_point(size = 5, aes(color = name %in% ROOTS)) +
    geom_node_text(aes(label = name), repel = TRUE, size = 3) +
    scale_color_manual(values = c("FALSE" = "grey70", "TRUE" = "lightblue"),
                       guide = "none") +
    labs(title = sprintf("%s — Averaged (t=%.3f)", NETWORK_ID, as.numeric(thresh_opt))) +
    theme_void()
}
```

# 8. Fit parameters and save

```{r fit-and-save}
# Fit parameters on the tabu DAG
fit_tabu <- bn.fit(bn_tabu, bn_df)

# Save everything
models <- list(
  bn_hc            = bn_hc,
  bn_tabu          = bn_tabu,
  bn_mmhc          = bn_mmhc,
  fit_tabu         = fit_tabu,
  boot_str         = boot_str,
  avg_opt          = avg_opt,
  avg_conservative = avg_conservative,
  avg_50           = avg_50,
  blacklist        = blacklist,
  config           = list(
    NETWORK_ID       = NETWORK_ID,
    SEED             = SEED,
    ROOTS            = ROOTS,
    NODE_MAP         = NODE_MAP,
    LOG_TRANSFORM    = LOG_TRANSFORM,
    REFERENCE_LEVELS = REFERENCE_LEVELS,
    MAXP             = MAXP
  )
)

out_path <- file.path("output", "models", sprintf("%s_models.rds", NETWORK_ID))
saveRDS(models, out_path)
cat(sprintf("Saved all models: %s\n", out_path))
```

# 9. Save DAG figures

```{r save-figures}
fig_dir <- file.path("output", "figures")

# Edge strength histogram
p <- ggplot(boot_str, aes(x = strength)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = c(0.5, 0.85, thresh_opt, thresh_conservative),
             linetype = c("dashed", "dashed", "solid", "solid"),
             color = c("orange", "red", "darkgreen", "darkgreen")) +
  annotate("text", x = 0.5, y = Inf, label = "0.50", vjust = 2, color = "orange") +
  annotate("text", x = 0.85, y = Inf, label = "0.85 (ref)", vjust = 2, color = "red") +
  annotate("text", x = thresh_opt, y = Inf, label = sprintf("opt=%.2f", thresh_opt),
           vjust = 3.5, color = "darkgreen") +
  annotate("text", x = thresh_conservative, y = Inf,
           label = sprintf("cons=%.2f", thresh_conservative),
           vjust = 5, color = "darkgreen") +
  labs(title = sprintf("%s — Bootstrap Edge Strength Distribution", NETWORK_ID),
       x = "Edge Strength", y = "Count") +
  theme_minimal()

ggsave(file.path(fig_dir, sprintf("%s_edge_strength_hist.png", NETWORK_ID)),
       p, width = 8, height = 5, dpi = 150)
cat("Saved edge strength histogram\n")
```
