---
title: "BN Structure Learning"
author: "BPGbio Capstone"
date: "`r Sys.Date()`"
output: html_document
params:
  config: "config/L1b_clinical_curated.R"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE,
                      fig.width = 10, fig.height = 10)
knitr::opts_knit$set(root.dir = here::here("bn_learning"))
library(bnlearn)
library(dplyr)
library(ggplot2)
```

# 1. Load config and data

```{r load}
source(params$config)
cat(sprintf("Network: %s\n", NETWORK_ID))

bn_df <- readRDS(file.path("output", "models", sprintf("%s_data.rds", NETWORK_ID)))
cat(sprintf("Data: %d samples x %d nodes\n", nrow(bn_df), ncol(bn_df)))

set.seed(SEED)
```

# 2. Generate blacklist

No edges into root nodes + roots mutually independent.
CLG constraint: discrete nodes cannot have continuous parents.

```{r blacklist}
all_nodes <- names(bn_df)
non_roots <- setdiff(all_nodes, ROOTS)

# Rule 1: No edges INTO root nodes (from any node)
bl_into_roots <- expand.grid(from = all_nodes, to = ROOTS, stringsAsFactors = FALSE) %>%
  filter(from != to)

# Rule 2: Roots mutually independent (no edges between roots in either direction)
# Already covered by Rule 1 (roots are in the "to" set), but make explicit:
bl_between_roots <- expand.grid(from = ROOTS, to = ROOTS, stringsAsFactors = FALSE) %>%
  filter(from != to)

blacklist <- rbind(bl_into_roots, bl_between_roots) %>% distinct()
cat(sprintf("Blacklist: %d edges forbidden\n", nrow(blacklist)))
```

# 3. Score-based structure learning

## 3a. Hill-climbing

```{r hc}
set.seed(SEED)
bn_hc <- hc(bn_df, score = "bic-cg", blacklist = blacklist)
cat(sprintf("HC: %d edges, BIC-cg = %.2f\n", narcs(bn_hc), score(bn_hc, bn_df, type = "bic-cg")))
```

## 3b. Tabu search

```{r tabu}
set.seed(SEED)
bn_tabu <- tabu(bn_df, score = "bic-cg", blacklist = blacklist)
cat(sprintf("Tabu: %d edges, BIC-cg = %.2f\n", narcs(bn_tabu), score(bn_tabu, bn_df, type = "bic-cg")))
```

## 3c. MMHC (hybrid)

```{r mmhc}
set.seed(SEED)
bn_mmhc <- mmhc(bn_df, blacklist = blacklist)
cat(sprintf("MMHC: %d edges, BIC-cg = %.2f\n", narcs(bn_mmhc), score(bn_mmhc, bn_df, type = "bic-cg")))
```

# 4. Algorithm comparison

```{r algorithm-comparison}
# Edge overlap
arcs_hc   <- apply(arcs(bn_hc), 1, paste, collapse = "->")
arcs_tabu <- apply(arcs(bn_tabu), 1, paste, collapse = "->")
arcs_mmhc <- apply(arcs(bn_mmhc), 1, paste, collapse = "->")

all_arcs <- unique(c(arcs_hc, arcs_tabu, arcs_mmhc))
comparison <- data.frame(
  edge = all_arcs,
  hc   = all_arcs %in% arcs_hc,
  tabu = all_arcs %in% arcs_tabu,
  mmhc = all_arcs %in% arcs_mmhc,
  stringsAsFactors = FALSE
) %>%
  mutate(n_agree = hc + tabu + mmhc) %>%
  arrange(desc(n_agree), edge)

cat(sprintf("Total unique edges across algorithms: %d\n", nrow(comparison)))
cat(sprintf("Edges in all 3: %d\n", sum(comparison$n_agree == 3)))
cat(sprintf("Edges in 2/3: %d\n", sum(comparison$n_agree == 2)))
cat(sprintf("Edges in 1/3: %d\n", sum(comparison$n_agree == 1)))

knitr::kable(comparison, caption = "Edge agreement across algorithms")
```

# 5. Bootstrap model averaging

```{r bootstrap, cache=TRUE}
set.seed(SEED)
boot_str <- boot.strength(
  bn_df,
  R = 1000,
  algorithm = "tabu",
  algorithm.args = list(score = "bic-cg", blacklist = blacklist)
)

cat("Bootstrap strength summary:\n")
print(summary(boot_str$strength))
cat(sprintf("\nDirection probability summary:\n"))
print(summary(boot_str$direction))
```

## 5a. Averaged networks at multiple thresholds

```{r averaged-networks}
# Optimal threshold (bnlearn heuristic)
avg_opt <- averaged.network(boot_str)
thresh_opt <- attr(avg_opt, "threshold")
cat(sprintf("Optimal threshold: %.4f\n", thresh_opt))
cat(sprintf("Averaged network (optimal, t=%.4f): %d edges\n", thresh_opt, narcs(avg_opt)))

# Conservative threshold (0.85)
avg_85 <- averaged.network(boot_str, threshold = 0.85)
cat(sprintf("Averaged network (t=0.85): %d edges\n", narcs(avg_85)))

# Liberal threshold (0.50)
avg_50 <- averaged.network(boot_str, threshold = 0.50)
cat(sprintf("Averaged network (t=0.50): %d edges\n", narcs(avg_50)))
```

# 6. Edge strength table

```{r edge-strength}
# Filter to edges with strength > 0.1 for readability
sig_edges <- boot_str %>%
  filter(strength > 0.1) %>%
  arrange(desc(strength))

knitr::kable(sig_edges, digits = 3,
             caption = "Bootstrap edge strengths (strength > 0.1)")
```

# 7. DAG visualizations

```{r plot-dag-tabu, fig.cap="Tabu search DAG"}
if (requireNamespace("Rgraphviz", quietly = TRUE)) {
  graphviz.plot(bn_tabu, main = sprintf("%s — Tabu", NETWORK_ID),
                shape = "ellipse",
                highlight = list(nodes = ROOTS, fill = "lightblue"))
} else {
  # Fallback: igraph/ggraph
  library(igraph)
  library(ggraph)
  g <- igraph::graph_from_data_frame(arcs(bn_tabu))
  ggraph(g, layout = "fr") +
    geom_edge_link(arrow = arrow(length = unit(3, "mm")), end_cap = circle(3, "mm")) +
    geom_node_point(size = 5, aes(color = name %in% ROOTS)) +
    geom_node_text(aes(label = name), repel = TRUE, size = 3) +
    scale_color_manual(values = c("FALSE" = "grey70", "TRUE" = "lightblue"),
                       guide = "none") +
    labs(title = sprintf("%s — Tabu", NETWORK_ID)) +
    theme_void()
}
```

```{r plot-dag-avg, fig.cap="Averaged network (optimal threshold)"}
if (requireNamespace("Rgraphviz", quietly = TRUE)) {
  graphviz.plot(avg_opt, main = sprintf("%s — Averaged (t=%.3f)", NETWORK_ID, thresh_opt),
                shape = "ellipse",
                highlight = list(nodes = ROOTS, fill = "lightblue"))
} else {
  library(igraph)
  library(ggraph)
  g <- igraph::graph_from_data_frame(arcs(avg_opt))
  ggraph(g, layout = "fr") +
    geom_edge_link(arrow = arrow(length = unit(3, "mm")), end_cap = circle(3, "mm")) +
    geom_node_point(size = 5, aes(color = name %in% ROOTS)) +
    geom_node_text(aes(label = name), repel = TRUE, size = 3) +
    scale_color_manual(values = c("FALSE" = "grey70", "TRUE" = "lightblue"),
                       guide = "none") +
    labs(title = sprintf("%s — Averaged (t=%.3f)", NETWORK_ID, thresh_opt)) +
    theme_void()
}
```

# 8. Fit parameters and save

```{r fit-and-save}
# Fit parameters on the tabu DAG
fit_tabu <- bn.fit(bn_tabu, bn_df)

# Save everything
models <- list(
  bn_hc     = bn_hc,
  bn_tabu   = bn_tabu,
  bn_mmhc   = bn_mmhc,
  fit_tabu  = fit_tabu,
  boot_str  = boot_str,
  avg_opt   = avg_opt,
  avg_85    = avg_85,
  avg_50    = avg_50,
  blacklist = blacklist,
  config    = list(
    NETWORK_ID       = NETWORK_ID,
    SEED             = SEED,
    ROOTS            = ROOTS,
    NODE_MAP         = NODE_MAP,
    LOG_TRANSFORM    = LOG_TRANSFORM,
    REFERENCE_LEVELS = REFERENCE_LEVELS
  )
)

out_path <- file.path("output", "models", sprintf("%s_models.rds", NETWORK_ID))
saveRDS(models, out_path)
cat(sprintf("Saved all models: %s\n", out_path))
```

# 9. Save DAG figures

```{r save-figures}
fig_dir <- file.path("output", "figures")

# Edge strength histogram
p <- ggplot(boot_str, aes(x = strength)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = c(0.5, 0.85, thresh_opt),
             linetype = c("dashed", "dashed", "solid"),
             color = c("orange", "red", "darkgreen")) +
  annotate("text", x = 0.5, y = Inf, label = "0.50", vjust = 2, color = "orange") +
  annotate("text", x = 0.85, y = Inf, label = "0.85", vjust = 2, color = "red") +
  annotate("text", x = thresh_opt, y = Inf, label = sprintf("opt=%.2f", thresh_opt),
           vjust = 3.5, color = "darkgreen") +
  labs(title = sprintf("%s — Bootstrap Edge Strength Distribution", NETWORK_ID),
       x = "Edge Strength", y = "Count") +
  theme_minimal()

ggsave(file.path(fig_dir, sprintf("%s_edge_strength_hist.png", NETWORK_ID)),
       p, width = 8, height = 5, dpi = 150)
cat("Saved edge strength histogram\n")
```
